Visualising MSAs

### 2022-03-16 analysis/MSA.Rmd
fix MSA code to be nicer

```{r fig.width=12}
ggmsa('data/OG0000189.RiceAraSeagrasses.aln.fasta', start = 221, end = 260, char_width = 0.5, seq_name = T) + geom_seqlogo() + geom_msaBar()
```

@@

x <- readAAStringSet('data/OG0000189.RiceAraSeagrasses.aln.fasta')
d <- as.dist(stringDist(x, method = "hamming")/width(x)[1])
tree <- bionj(d)
p <- ggtree(tree) + geom_tiplab()
ggtree(tree) + geom_tiplab()
```

```{r fig.width=10}
data <- tidy_msa('data/OG0000189.RiceAraSeagrasses.aln.fasta', 221, 260)
p + geom_facet(geom = geom_msa, data = data,  panel = 'Multiple Sequence Alignment (221-260AA)',
               font = NULL, color = "Chemistry_AA") +
    xlim_tree(1)
```

@@

```{r}
treestring <- '((3O.sativa:0.11386470583955717040,4O.sativa:0.03615896391829190315):0.07379737996482453599,(5O.sativa:0.33751444720146173140,((7A.thalian:0.45696796911334131019,((8Z.mueller:0.33996861614583556710,15Z.marina:0.16088168380872580610):0.31923212677234219514,10P.austra:0.57486248434538900209):0.22324965676023655892):0.10039129162924786964,((13Z.marina:0.11346816437532522559,(12Z.muelle:0.00973855109814669891,10Z.muelle:0.02645894329604922546):0.09840311662577182206):0.35332170475064506032,((14Z.marina:0.12821969530885793387,(9Z.mueller:0.06883167458475099310,11Z.muelle:0.02355162378347611801):0.15839770535961147924):0.63099939044074537797,(1A.amphibo:0.53760770644490452064,6P.austral:0.07001018412366488697):0.08839123570898912985):0.05074368132159561007):0.09141392275039983417):0.12967268161775510893):0.11320488597472283532,2O.sativa:1.70657194075952300949):0.0;'
plot(ape::read.tree(text=treestring))
p2 <- ggtree(ape::read.tree(text=treestring)) + geom_tiplab()

@@

Let's rename the tree and the MSA data table using `rename_taxa`
```{r}
trees <- ape::read.tree(text=treestring)
old_names <- trees$tip.label
# [1] "3O.sativa"  "4O.sativa"  "5O.sativa"  "7A.thalian" "8Z.mueller" "15Z.marina" "10P.austra" "13Z.marina"
# [9] "12Z.muelle" "10Z.muelle" "14Z.marina" "9Z.mueller" "11Z.muelle" "1A.amphibo" "6P.austral" "2O.sativa"
new_names  <- c( 'OsCBL8', 'OsCBL7.1', 'OsCBL4', 'AtSOS3', 'ZmuSOS3', 'ZmaSOS3', 'PaSOS3', 'ZmaSOS3.2', 'ZmuSOS3.2', 'ZmuSOS3.3', 'ZmaSOS3.2', 'ZmuSOS3.4', 'ZmuSOS3.5', 'AaSOS3', 'PaSOS3.2', 'OsCBL7.2')
rename_df <- data.frame(old = old_names, new = new_names)
rename_df

@@

```{r}
#str_replace_all takes a named vector
replace_vector <- new_names
names(replace_vector) <- old_names
data2 <- data2 %>% mutate(name = str_replace_all(name, replace_vector))
```

@@

```
 # to get best model
modeltest-ng -i OG0000189.RiceAraSeagrasses.aln.fasta -t ml -d aa -p 8

@@

raxml-ng --support --tree T3.raxml.bestTree --bs-trees allbootstraps --prefix T13 --redo
```

@@

data <- tidy_msa('data/OG0000189.RiceAraSeagrasses.aln.fasta', 221, 260)
data3 <- data %>% mutate(name = str_replace(name, ' ', '_'),
                          name = str_replace_all(name, replace_vector))
p3 + geom_facet(geom = geom_msa, data = data3,  panel = 'Multiple Sequence Alignment (221-260AA)',
               font = NULL, color = "Chemistry_AA") +
    xlim_tree(3)
```

>>>>

```{r fig.width=12}
ggmsa('data/SOS3_OG0000189/OG0000189.RiceAraSeagrasses.aln.fasta', start = 221, end = 260, char_width = 0.5, seq_name = T) + geom_seqlogo() + geom_msaBar()
```

@@

x <- readAAStringSet('data/SOS3_OG0000189/OG0000189.RiceAraSeagrasses.aln.fasta')
d <- as.dist(stringDist(x, method = "hamming")/width(x)[1])
tree <- bionj(d)
p <- ggtree(tree) + geom_tiplab()
ggtree(tree) + geom_tiplab()
```

```{r fig.width=10}
data <- tidy_msa('data/SOS3_OG0000189/OG0000189.RiceAraSeagrasses.aln.fasta', 221, 260)
p + geom_facet(geom = geom_msa, data = data,  panel = 'Multiple Sequence Alignment (221-260AA)',
               font = NULL, color = "Chemistry_AA") +
    xlim_tree(1)
```

@@

```{r}
treestring <- '((3O.sativa:0.11386470583955717040,4O.sativa:0.03615896391829190315):0.07379737996482453599,(5O.sativa:0.33751444720146173140,((7A.thalian:0.45696796911334131019,((8Z.mueller:0.33996861614583556710,15Z.marina:0.16088168380872580610):0.31923212677234219514,10P.austra:0.57486248434538900209):0.22324965676023655892):0.10039129162924786964,((13Z.marina:0.11346816437532522559,(12Z.muelle:0.00973855109814669891,10Z.muelle:0.02645894329604922546):0.09840311662577182206):0.35332170475064506032,((14Z.marina:0.12821969530885793387,(9Z.mueller:0.06883167458475099310,11Z.muelle:0.02355162378347611801):0.15839770535961147924):0.63099939044074537797,(1A.amphibo:0.53760770644490452064,6P.austral:0.07001018412366488697):0.08839123570898912985):0.05074368132159561007):0.09141392275039983417):0.12967268161775510893):0.11320488597472283532,2O.sativa:1.70657194075952300949):0.0;'
plot(ape::read.tree(text=treestring))
p2 <- ggtree(ape::read.tree(text=treestring)) + geom_tiplab()
p2
```

@@

Let's rename the tree and the MSA data table using `rename_taxa`
```{r}
trees <- ape::read.tree(text=treestring)

rename_df <- tibble::tribble(
                       ~old,        ~new,
                "3O.sativa",    "OsCBL8",
                "4O.sativa",  "OsCBL7.1",
                "5O.sativa",    "OsCBL4",
               "7A.thalian",    "AtSOS3",
               "8Z.mueller",   "ZmuSOS3",
               "15Z.marina",   "ZmaSOS3",
               "10P.austra",    "PaSOS3",
               "13Z.marina", "ZmaSOS3.2",
               "12Z.muelle", "ZmuSOS3.2",
               "10Z.muelle", "ZmuSOS3.3",
               "14Z.marina", "ZmaSOS3.2",
               "9Z.mueller", "ZmuSOS3.4",
               "11Z.muelle", "ZmuSOS3.5",
               "1A.amphibo",    "AaSOS3",
               "6P.austral",  "PaSOS3.2",
                "2O.sativa",  "OsCBL7.2"
               )

@@

```{r}
#str_replace_all takes a named vector
replace_vector <- rename_df$new
names(replace_vector) <- rename_df$old
data2 <- data2 %>% mutate(name = str_replace_all(name, replace_vector))
```

@@

## Phylogeny for the SOS3 cluster
```

 # to get best model
modeltest-ng -i OG0000189.RiceAraSeagrasses.aln.fasta -t ml -d aa -p 8

@@

raxml-ng --support --tree T3.raxml.bestTree --bs-trees allbootstraps --prefix T13

# let's make another 400 so we can have a nice 1000 bootstraps - different seed again!

raxml-ng --msa OG0000189.RiceAraSeagrasses.aln.fasta.raxml.reduced.phy --model JTT-DCMUT+G4 --prefix T14 --threads 8 --seed 1234 --bs-trees 400  --bootstrap

cat T8.raxml.bootstraps T11.raxml.bootstraps T14.raxml.bootstraps > allbootstraps
raxml-ng --support --tree T3.raxml.bestTree --bs-trees allbootstraps --prefix T15

@@

data <- tidy_msa('data/SOS3_OG0000189/OG0000189.RiceAraSeagrasses.aln.fasta', 221, 260)
data3 <- data %>% mutate(name = str_replace(name, ' ', '_'),
                          name = str_replace_all(name, replace_vector))
p3 + geom_facet(geom = geom_msa, data = data3,  panel = 'Multiple Sequence Alignment (221-260AA)',
               font = NULL, color = "Chemistry_AA") +
    xlim_tree(3)
```
## Phylogeny for the EIN3 cluster

First, to align the sequences again:

```
modeltest-ng reports JTT+I+G4+F

raxml-ng --msa OG0000629.RiceAraSeagrasses.aln.fa --model JTT+I+G4+F

 # to get fixed fasta - all good!!
raxml-ng --msa OG0000629.RiceAraSeagrasses.aln.fa --model JTT+I+G4+F --check

# to make a regular tree
raxml-ng --msa OG0000629.RiceAraSeagrasses.aln.fa --model JTT+I+G4+F --prefix T3 --threads 2 --seed 2

# make 1000 bootstrap trees
raxml-ng --msa OG0000629.RiceAraSeagrasses.aln.fa --model JTT+I+G4+F --prefix T8 --threads 16 --seed 2 --bootstrap --bs-trees 1000

# check whether they converge with <3% WRF cutoff
raxml-ng --bsconverge --bs-trees T8.raxml.bootstraps --prefix T12 --seed 2 --threads 1 --bs-cutoff 0.03

# it actually converged with 50 trees. haha.

raxml-ng --support --tree T3.raxml.bestTree --bs-trees allbootstraps --prefix T13 
```

There we go:

```{r}
tree3 <- ape::read.tree(text='(((LOC_Os07g48630.1:0.641123,(LOC_Os03g20780.1:0.007239,LOC_Os03g20790.1:0.001048)100:0.214718)100:0.418001,(((Zosma02g24210:0.119273,(maker-1834_92013--0.52-mRNA-1:0.074816,augustus_masked-5412_21708--0.0-mRNA-1:0.082668)97:0.073791)100:0.640955,augustus_masked-scf7180003860583-processed-gene-0.3-mRNA-1:0.182720)89:0.114788,(maker-3356_51779--0.18-mRNA-1:0.661393,Zosma01g42670:0.320772)100:0.468870)67:0.098845)100:0.660268,AT3G20770.1:0.149857,AT2G27050.1:0.250165)100:0.0;', branch.label='support')

rename_df <- tibble::tribble(
                                                       ~old,        ~new,
                                         "LOC_Os07g48630.1",    "OsEIL2",
                                         "LOC_Os03g20780.1",    "OsEIL1",
                                         "LOC_Os03g20790.1",  "OsEIL1.2",
                                            "Zosma02g24210", "ZmaEIN3.1",
                            "maker-1834_92013--0.52-mRNA-1", "ZmuEIN3.1",
                   "augustus_masked-5412_21708--0.0-mRNA-1", "ZmuEIN3.2",
"augustus_masked-scf7180003860583-processed-gene-0.3-mRNA-1",   "PaEIN3",
                            "maker-3356_51779--0.18-mRNA-1", "ZmuEIN3.3",
                                            "Zosma01g42670", "ZmaEIN3.2",
                                              "AT3G20770.1",    "AtEIN3",
                                              "AT2G27050.1",    "AtEIL1"
               )

tree3 <- rename_taxa(tree3, rename_df, old, new)

p3 <-  ggtree(tree3) + geom_tiplab(fontface='italic')
p3 <- p3 + geom_nodelab(aes(subset=label>80))
p3 + xlim_tree(2.1)
```
